Galois Characterization of Function Fields of \\ Transcendence Degree $1$ over Number Fields

### 2017-05-10 Master Thesis/synopsis.tex
Add some chapter comments

@@

\begin{lemma}\label{2.2-pre}
For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
\begin{align*}
\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} &\longrightarrow \operatorname{Spv}k(v)\\ 
w&\longmapsto w\mod v
\end{align*}
Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e. $k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$. In other words, $v$ is a closed point iff $\dim k(v)=0$. By the dimension inequality \ref{rank-dimension-inequality} we have $\operatorname{Spv}^{(n)}(K)\subset |\operatorname{Spv}(K)|$ where $n=\dim(K)$.

If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{d}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$. $w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
\end{lemma}

@@

\begin{lemma}\phantomsection\label{lemma-defectless-valuations}
\begin{enumerate}[(i)]
\item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)$
\end{enumerate}
\end{lemma}

\begin{proof}
\begin{enumerate}[(i)]
\item Let $v$ be a defectless valuation of rational rank $1$. By Lemma \ref{rank-dimension-inequality}, $\operatorname{rank}(v) \leq \operatorname{rrank}(v) = 1$. Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality. \qedhere
\end{enumerate}
\end{proof}

@@

@@

@@

@@

@@

\begin{proof}
Let $T$ be the absolute inertia group of $K$ and consider the short exact sequence from Theorem \ref{theorem-absoluteinertia}:
\[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]
Let $G_\ell$ be an $\ell$-Sylow group of $G_K$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$. Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial. The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$. However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By [Se], Ch. I, §4, Proposition 22 we have:
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of [Se], Ch. I, §4, Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

@@

@@

@@

@@

@@

\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that now the following diagram commutes: 
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]

>>>>

This section we recall facts from valuation theory.

@@

\begin{lemma}\label{2.2-pre}
For a fixed $v\in \operatorname{Spv}(K)$ there is a canonical bijection: 
\begin{align*}
\{ w\in\operatorname{Spv}(K)\mid \mathcal{O}_w\subset\mathcal{O}_v \} &\longrightarrow \operatorname{Spv}k(v)\\ 
w&\longmapsto w\mod v
\end{align*}
Therefore $v$ is a closed point iff $k(v)$ only admits the trivial valuation, i.e. $k(v)$ has characteristic $p>0$ and is algebraic over $\mathbb{F}_p$. In other words, $v$ is a closed point iff $\dim k(v)=0$.

If $w\in\operatorname{Spv}_\text{rd}^{(i)}(K)$ for some $i\geq 1$ then there exists a unique $v(w)\in\operatorname{Spv}_\text{rd}^{(i-1)}(K)$ such that $\mathcal{O}_w\subset\mathcal{O}_{v(w)}$. $w$ is henselian if and only if $v(w)$ and $w\mod v(w)$ are henselian.
\end{lemma}

@@

\begin{lemma}\phantomsection\label{lemma-defectless-valuations}
\begin{enumerate}[(i)]
\item $\operatorname{Spv}_\text{d}^{(1)}(K) = \operatorname{Spv}_\text{rd}^{(1)}(K)$
\item $\operatorname{Spv}^{(2)}(K) = \operatorname{Spv}_\text{d}^{(2)}(K)\subset |\operatorname{Spv}(K)|$
\end{enumerate}
\end{lemma}

\begin{proof}
\begin{enumerate}[(i)]
\item Let $v$ be a defectless valuation of rational rank $1$. By Lemma \ref{rank-dimension-inequality}, $\operatorname{rank}(v) \leq \operatorname{rrank}(v) = 1$. Since $\Gamma_v$ is not trivial, $\operatorname{rank}(v) = 1=\operatorname{rrank}(v)$.
\item Let $v$ be a valuation of rational rank $2$. Due to Lemma \ref{rank-dimension-inequality}, the dimension inequality is an equality and we have $\dim k(v) = 0$. \qedhere
\end{enumerate}
\end{proof}

@@

The following lemma is the key result in proving Theorem \ref{thm:main-result}. First, we use K. Kato's two-dimensional local-global principle to show the assertion for finitely generated $K$. Then, we'll use a quasi-compactness argument to see that we can drop the assumption that $K$ is finitely generated.

@@

In order to generalize Lemma \ref{2.4-finitelygenerated} for arbitrary, not necessarily finitely generated $K$, we need the following Lemma:

@@

Now the general case follows by a quasi-compactness argument:

@@

Now we calculate the cohomology groups $\mathrm{H}^3(K_w,\mathbb{Z}/\ell\mathbb{Z})$ for $w\in |\operatorname{Spv}(K)|$. With the help of Lemma \ref{2.4} we can find a cohomological criterion of henselian fields over $K$ with respect to a $w\in\operatorname{Spv}_\text{d}^{(2)}(K)$.

@@

\begin{proof}
Let $T$ be the absolute inertia group of $K$ and consider the short exact sequence from Theorem \ref{theorem-absoluteinertia}:
\[ 1\longrightarrow T\longrightarrow G_K\longrightarrow G_k\longrightarrow 1 \]

Let $G_\ell$ be an $\ell$-Sylow group of $G_K$. By Theorem \ref{theorem-sylowgroups-inertiagroup}, the $\ell$-Sylow group of $T$ isomorphic to $\mathbb{Z}_\ell^r$, where $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)$. Furthermore, $G_k\subset G_{\mathbb{F}_p}\cong\hat{\mathbb{Z}}$, thus the $\ell$-Sylow group $G_{k,\ell}$ of $G_k$ is either isomorphic to $\mathbb{Z}_\ell$ or trivial. The exact sequence above gives rise to the short exact sequence of their $\ell$-Sylow subgroups:
\[ 0\longrightarrow\mathbb{Z}_\ell^r \longrightarrow G_\ell \longrightarrow G_{k,\ell}\longrightarrow 1 \]
Since $G_{k,\ell}$ is a free pro-$\ell$ group, this sequence splits and we have an isomorphism $G_\ell \cong \mathbb{Z}_\ell^r\rtimes G_{k,\ell}$ with $r =\dim_{\mathbb{F}_\ell}(\Gamma/\ell\Gamma)\leq \operatorname{rrank}(\Gamma)\leq 2$.

Let $G_K(\ell)$ be the maximal pro-$\ell$ quotient of $G_K$. Since $\mu_\ell\subset K$, the action of $G_k$ on the $\ell$-Sylow group of the inertia group of $w$ factors through the maximal pro-$\ell$ quotient $G_k(\ell)$ of $G_k$. Since $G_{k,\ell}\cong G_k(\ell)$, we also see that the canonical projection $G_K\to G_K(\ell)$ restricts to an isomorphism $G_\ell\to G_K(\ell)$. Now consider the following composition:
\[ \begin{tikzcd}
\mathrm{H}^3(G_K(\ell),\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{inf}"] & \mathrm{H}^3(G_K,\mathbb{Z}/\ell\mathbb{Z}) \ar[r, "\operatorname{res}"] & \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})
\end{tikzcd} \]
Since $\operatorname{res}\circ\operatorname{inf}$ is an isomorphism, $\operatorname{inf}$ is injective and $\operatorname{res}$ is surjective. We have $\operatorname{cor}\circ\operatorname{res} = [G_K:G_\ell]$. However, $[G_K:G_\ell]$ is coprime to $\ell$, which makes $\operatorname{res}$ injective as well. Therefore, $\operatorname{inf}$ and $\operatorname{res}$ are both isomorphisms and it suffices to calculate $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})$. By \cite{Se64}, Ch. I, §4, Proposition 22 we have:
\[\operatorname{cd}(G_\ell) = \operatorname{cd}(\mathbb{Z}_\ell^r\rtimes G_k(\ell)) = \begin{cases}
r, & G_k(\ell) = 0 \\
r + 1, & G_k(\ell)\cong\mathbb{Z}_\ell
\end{cases} \] 
since $\operatorname{cd}(\mathbb{Z}_\ell)=1$. Thus $\mathrm{H}^3(G_\ell, \mathbb{Z}/\ell\mathbb{Z}) =0$ if $G_k(\ell) = 0$ or $r < 2$. Now let $G_k(\ell) \cong\mathbb{Z}_\ell$ and $r = 2$. According to the proof of \cite{Se64}, Ch. I, §4, Proposition 22, we have:
\[ \mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z}) = \mathrm{H}^1(G_k(\ell), \mathrm{H}^2 (\mathbb{Z}_\ell^r, \mathbb{Z}/\ell\mathbb{Z})) \]
By [NSW] Theorem 2.4.6, we have $\mathrm{H}^2(\mathbb{Z}_\ell^r,\mathbb{Z}/\ell\mathbb{Z}) \cong\mathrm{H}^1(\mathbb{Z}_\ell, \mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})) = \mathbb{Z}/\ell\mathbb{Z}$, since $\mathrm{H}^1(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\operatorname{Hom}_\text{cont.}(\mathbb{Z}_\ell, \mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$. Therefore, $\mathrm{H}^3(G_\ell,\mathbb{Z}/\ell\mathbb{Z})=\mathbb{Z}/\ell\mathbb{Z}$.

@@

In this section we want to see that $\sigma$ induces bijections on defectless valuations. A direct consequence of Proposition \ref{2.6} is the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$:

@@

Using the bijection $\operatorname{Spv}_\text{d}^{(2)}(K)\to \operatorname{Spv}_\text{d}^{(2)}(K')$ above we want to establish a bijection $\operatorname{Spv}_\text{rd}^{(2)}(K)\to \operatorname{Spv}_\text{rd}^{(2)}(K')$ in order to establish a bijection between the defectless rank-$1$ valuations:

@@

In this section we shall show that $\sigma: G_K\to G_{K'}$ induces group isomorphisms $K'^\times/n\to K^\times/n$ for all $n\in\mathbb{N}$.

@@

First of all, we want to show that $\sigma$ induces an isomorphism between the absolute Galois groups of the constant fields $G_k$ and $G_{k'}$. We need the following Lemma:

@@

\paragraph{} Let $n\in\mathbb{N}$. In the previous remark, we have replaced $\sigma$, so that the following diagram commutes now:
\[ \begin{tikzcd}
G_K \ar[r]\ar[d, "\sigma"'] & G_k\\
G_{K'}\ar[ru]
\end{tikzcd} \]
The operation of $G_K$ and $G_{K'}$ on $\mu_n$ factors through $G_k$, thus we have $\sigma(g)\xi = g\xi$ for all $\xi\in\mu_n$ and $g\in G_K$. Therefore, for all intermediate fields $L$ of $\overline{K}/K$, $\sigma$ induces an isomorphism:
\[ \varphi_n:L'^\times/n \cong\mathrm{H}^1(L', \mu_n) \to \mathrm{H}^1(L,\mu_n) \cong L^\times/n \]
